BSTERGMs: Bayesian Separable-Temporal Exponential family Graph Models

### 2020-12-30 doc/paper_format.tex
revise doc

Nowadays, new kinds of data having special forms emerge in many fields.
A series of real number values do not confine these new data as traditional data are in the past.
The data may have a non-number type of value, unique structure, or even integrated different types of values with model-specific relation.
These data are called complex data, and the attention of statistical analysis on the data increases over time.

As one form of complex data, networks, or random graphs get highlighted attention among complex data for statistical analysis.
A relational structure can represent many phenomena in the real world, and the structure can be expressed effectively by a form of a network. 
For example, social networks, including offline and online, can be thought of as networks.
In this view, some social aspects of the relationship between actors can be modeled under an object of social science, 
including an analysis of power structure and a summary of intimacy structure on a given group.
Not only that, application to a transmitted disease in epidemiology enables statistical modeling based on real data.

The ability to capture distinct features on the observed network enlarges statistical models' usefulness on networks. 
The statistical models give information about local relations like formation or dissolution between two actors 
as well as global features of the whole network, such as the probability that a specific structure realizes. 
The statistical models investigate the complex behavior of networks, 
including how an observed network's global features may be related to local network structure. 
During the investigation, the statistical models can capture existing dependency on a given network and provide a tool 
to understand what certain structural features are more common on rare.

Exponential-family random graph models(ERGMs) may be the most useful tools to analyze cross-sectional networks and their dependencies.
\cite{RN117}, \cite{RN123}, \cite{RN124} and \cite{RN122} proposed the models and apply them to analyze dependent structures, 
and \cite{RN115} and \cite{RN116} started the Bayesian approaches of the models.
In the ERGMs, The dependencies are captured using a series of local features called network statistics from classical graph-theory. 
Since these models play an important role throughout this paper, I will introduce them more precisely in the following chapter.

The next natural question is a way to model a network sequence or network evolution over time. 
For instance, one may want to model social network structures for one year or 
simulate a disease infection based on the sequential observations in the past. 
The first exploration applying ERGMs to a network sequence was \cite{RN109}. 
After them, a notable contributions were of \cite{RN125} and \cite{RN126}, the Temporal ERGMs(TERGMs). 
In the TERGMs, ERGMs are used by modeling the transition probability between times. 
Moreover, \cite{RN93} extend the TERGMs to the Separable Temporal ERGMs(STERGMs), 
considering the separation between duration - how long ties tend to last - and 
incidence - the rate at which new ties are formed -from 
prevalence - for example, the total number of ties - of the network properties. 
By imposing a conditional independence condition between the duration and incidence on the model, 
the models can be extended to more effectively interpretable ones using similar fitting methods with TERGMs 
without adding much complexity.

Although the STERGMs give us strength when interpreting the fitted parameter, 
a Bayesian approach for the STERGMs has not been conducted. 
It is partly because of the hardship from the massive computation in a parameter estimation method for the models. 
Nonetheless, the Bayesian approach of STERGMs provides the statistical inference method following the standard Bayesian way, 
so the approach has a huge advantage of convenience for statistical works for fitted parameters and predictions to next time network structure. Thus, in this paper, I specify the bayesian separable temporal exponential random graph models(BSTERGMs) and propose a fitting algorithm for the model parameter. I also discuss ways of inference, the diagnostics procedure, and checking the goodness of fit.

Chapter 2 reviews the background materials, including the ERGMs, the BERGMs, the TERGMs, and the STERGMs. 
In chapter 3, I propose the BSTERGMs and show the detailed setting of BSTERGMs. 
In the following chapter, chapter 4, I offer the estimation method for the BSTERGMs. 
Chapter 5 explains ways to conduct diagnostics, statistical inference, and goodness of fit checks. 
Besides, I illustrate an application of BSTERGMs to network data of scholars and faculty
from the UCLA medical working project team in chapter 6, 
Finally, I discuss some possible extensions and ways to supplement the BSTERGMs further for future research.

@@

\chapter{Background} \label{Chapter2}
\section{Random graphs}
Networks or random graphs play a central role as a general model 
describing relations among subjects in the statistical context. 
To begin with, I will briefly summarize some terminologies and notations for concepts related to the networks.

I will use two words, 'graphs' and 'network,' as the same meaning without distinction. 
We say each subject on the data to a node. 
They sometimes are called vertex in graph theory. 
In addition, we say a tie joining two nodes to an edge. A tie can be directed or not. 
These two components have a fundamental role in describing the relational structure. 
Define a graph or a network to $y=(V,E)$, where $V$ is a set of nodes, and $E$ is a set of edges of a network.

For simplicity, I will deal with a situation without weight on each node or edge.
Then, a graph $y$ can be represented by a more straightforward form.
Assume that the number of node $n\in \mathbb{N}$ on data is given.
Then, for $\mathcal{B}=\{0,1\}$,
the set $\mathcal{Y} \subset \mathcal{B}^{n^2}$ is a set of graphs of $n$ nodes.
Now, a graph $y \in \mathcal{Y}$ can be represented by a $n \times n$ binary matrix
with elements $y_{ij}=1$ if there is a tie from $i$-th node to $j$-th node,
otherwise $y_{ij}=0$.

@@

I extend the concept of graphs to describe a situation that connections among nodes are random.
Let $\Omega$ be an event set of all possible connected situations among given nodes.
We say $Y: \Omega \to \mathcal{Y}$ is a random variable for a graph, or a random graph.
For a random graph $Y \in \mathcal{Y}$, denote the edge between i-th node and j-th node by $Y_{ij}$ for $i,j=1,2,...,n$
satisfying $Y_{ij}=1$ if the edge is connected. Otherwise, $Y_{ij}=0$.
Note that the $Y_{ij}$s are also random variables.

@@

\section{ERGMs: Exponential random graph models}
One of the natural questions from the definition of the random graph is 
how we can model a probability distribution on $\mathcal{Y}$.
As a answer, the Exponential random graph models (ERGMs) (\cite{RN123} and \cite{RN122})
gives a classical but flexible scheme to model for time-invariant networks represented as a single directed or undirected graph.
Moreover, ERGMs provide a statistical approach to the systematic exploration of several local features simultaneously and 
how these local structures interact to form a global network.

Specifically, following the notation of the previous section,
let $\mathcal{Y}$ be a set of random graphs with $n$ nodes and $Y \in \mathcal{Y}$ be a random graph.
Set a distribution on $\mathcal{Y}$ to
    \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
for some $\theta\in\mathbb{R}^p$,
where $s(y)\in\mathbb{R}^p$ is a vector which is part of y's sufficient network statistics,
and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
Models that have such a form are called the ERGMs.

@@

\section{TERGMs: Temporal Exponential family Random Graphs Models}
Let's consider a sequence of random graph $\{Y_1, Y_2, ..., Y_T\} (T\in\mathbb{N})$ and assume that
we are interested in the dynamics or evolution of the network sequence.
The temporal Temporal Exponential family Random Graphs Models(TERGMs)
proposed by \cite{RN125} and \cite{RN126}
is a model for our interest if we permit a Markov assumption
\[P(Y_2,Y_3,...,Y_T|Y_1)=P(Y_2|Y_1)P(Y_3|Y_2)...P(Y_T|Y_{T-1})\]
and focus on the transition probability along index of the sequence.
For convenience, I refer to the index as a time.

Specifically, let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs in $\mathcal{Y}$.
Set a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds) to
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta) = \frac{exp(\theta^{T}s(y_t, y_{t-1}))}{c(\theta, y_{t-1})}\]
where $t=2,...,T$, $\theta\in\mathbb{R}^p$,
sufficient network statistics $s(y_t, y_{t-1})\in\mathbb{R}^n$,
and a normalizing constant $c(\theta, y_{t-1})=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y, y_{t-1}))$.
Models that have such a form are called the TERGMs.

@@

\section{STERGMs: Separable-Temporal Exponential family Random Graphs Models}
One solution to remedy the problems of TERGMs is 
the Separable-Temporal Exponential family Random Graphs Models, or simply STERGMs. %ref
The key idea is to separate the formation models and dissolution model
for decoupling an effect on edge's duration of the model parameter from edge's incidence
when setting a distribution on $\mathcal{Y}\times ... \times \mathcal{Y}$ ($T-1$ folds, $T\in\mathbb{N}$),
where $\mathcal{Y}$ is a set of graphs with given $n$ nodes.

Precisely,
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T$ be random graphs (on $\mathcal{Y}$).
Let $\mathcal{Y}^+|_t$ be a subset of $\mathcal{Y}$ consisting of all graphs which have equal or additional edges comparing to $y_{t-1}$.
Likewise, let $\mathcal{Y}^-|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or sparse edges comparing to $y_{t-1}$.
Next, for $Y_t^+: \Omega \to\mathcal{Y}^+|_t$, $Y_t^-: \Omega \to\mathcal{Y}^-|_t$, $y_t^+ \in \mathcal{Y}^+|_t$ and $y_t^- \in \mathcal{Y}^-|_t$, set
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for some $\theta^+,\theta^-\in\mathbb{R}^p$, $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$, which are parts of sufficient network statistics,
and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))$, 
and $c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

Then, defining operations $+,-$ on $\mathcal{Y}$ following the edgewise boolean algebra,
(For detail, Define binary operations on the edge values $\{0,1\}$ by $+$ and $-$ 
by $1+1=1, 1+0=0+1=1, 0+0=0$ and $1-1=0, 1-0=1, 0-1=0, 0-0=0$) 
set the combined network $y_t$ to
\[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

Additionally, assume the first-order Markov assumption,
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$, called the separability.
Then we get
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\]
Models that have such a form are called the STERGMs.

\section{Network Statistics for ERGMs and derived models}
Now, I briefly discuss the terms of network statistics.
As a usual, $s(y)$ of ERGMs or $s(y_{t}^.,y_{t-1})$ of TERGMs and STERGMs would 
reflect the structure of given network without identifying each node index.
Typical candidates of the statistics are here.

@@

Following \cite{RN104}, usual candidates are distributions defined by some sense of the structure of a given graph. 
For a given $i, 1\leq i \leq n-1$, $D_i(y)$ is defined as the number of nodes in y whose the number of edges incident to the node equals $i$, 
and called $i$-th order node degree distribution.

For a given $i, 1\leq i \leq n-2$, $P_i(y)$ is defined to be the number of dyads $(j,k), 1\leq j,k \leq n$
in y whose $j$ and $k$ have nodes connected by an edge each other and they share exactly $i$ connected nodes in common,
and called $i$-th order edgewise shared partner distribution.

Note that, the bundle of all order of node degree distribution and all order of edgewise shared partner distribution,
\[(D_1(y),D_2(y),...,D_{n-1}(y),P_1(y),P_2(y),...,P_{n-2}(y))\] 
forms sufficient statistics of a graph.
And there are relation that
\[n=D_0(y)+D_1(y)+...+D_{n-1}(y)\]
and
\[|y|=P_0(y)+P_1(y)+...+P_{n-2}(y)=\frac{1}{2}\sum_{i=1}^{n-1} iD_i(y)\]

A function of the sufficient statistics can be candidates for the model statistics.
\cite{RN128} proposed graph statistics including $k$-star $S_k(y)$ for $1\leq k \leq n-1$ and $k$-triangles $T_k$ for $1\leq k \leq n-2$.
The $k$-star consists of a node together with a set of $k$ of its connected nodes.
The $k$-triangle consists of $k$ triangles that share one common edge.
They can be expressed by the definition of node degree distributions and edgewise shared partner distributions,
\[S_1 = |y|, 
S_k = \sum_{i=1}^{n-1} \binom{i}{k} D_i(y), k\geq 2\]
and
\[T_1 = \frac{1}{3}\sum_{i=1}^{n-2} iP_i(y),
T_k = \sum_{i=1}^{n-2} \binom{i}{k} P_i(y), k\geq 2\]

Another interesting statistics are geometrically weighted value of distributions defined by
\cite{RN118}, \cite{RN104}, \cite{RN128} and \cite{RN103}.
For example, geometrically weighted (node) degree with parameter $\tau$ is defined by
\[GWD(y;\tau)=e^{\tau} \sum_{k=1}^{n-2} (1-(1-e^{-\tau})^k)D_k(y)\]
and geometrically weighted shared partner distribution with parameter $\tau$ is defined by
\[GWESP(y;\tau)=e^{\tau} \sum_{k=1}^{n-2} (1-(1-e^{-\tau})^k)P_k(y)\]

In the context of temporal models, parameters in the model take two network as a arguments like $s(y_t,y_{t-1})$.
One of the common choices for these statistics is the bundle of sufficient statistics of first argument $y_t$.
Another common choice is using a difference, $s(y_t,y_{t-1})=s'(y_t)-s'(y_{t-1})$ where $s'$ is a network statistics of given network.
Although there are many other statistics related to ERGMs, I omit them and end this section here.
For additional example, see \cite{RN107} for the ERGMs, and \cite{RN125}'s chapter 2.1 for temporal the ERGMs.

@@

The goal of this chapter is to convert the STERGMs to the Bayesian setting.
With priors $p(\theta^+),p(\theta^-)$ over $\theta^+,\theta^-$,
the Bayesian model for the STERGMs can be made. I call this model Bayesian Separable Temporal Exponential family Random Graph Models,
or the BSTERGMs.

More explicitly, for given the number of nodes $n$, let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T$ be random graphs (of $\mathcal{Y}$).
As the STERGMs, define $\mathcal{Y}^+|_t$, $\mathcal{Y}^-|_t$ be subsets of $\mathcal{Y}$ consisting all graphs which have equal or additional edges comparing to $y_{t-1}$,
and which have equal or sparse edges comparing to $y_{t-1}$ respectively.
Then, for $Y_t^+$ of $\mathcal{Y}^+|_t$ and $Y_t^-$ of $\mathcal{Y}^-|_t$, $y_t^+ \in \mathcal{Y}^+|_t$ and $y_t^- \in \mathcal{Y}^-|_t$, set
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for some $\theta^+,\theta^-\in\mathbb{R}^p$, $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^n$, which are parts of sufficient network statistics,
and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))$, 
and $c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

Then, with the edgewise boolean algebra, set the combined network $y_t$ to
\[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

and assume that the first-order Markov property
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and the conditional independence(separability) between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$.
Then we get
\[\theta^+ \sim p(\theta^+), \theta^- \sim p(\theta^-)\]
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-)=P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)\]

>>>>

Nowadays, new kinds of data with special forms emerge in many fields.
A sequence of real number values do not confine these new data as traditional data are in the past.
The data may have a non-number type of value, unique structure, or even integrated different types of values with model-specific relation.
These data are called complex data, and the attention of statistical analysis on the data increases over time.

As one form of complex data, network, or random graph gets highlighted attention as a flexible modeling tool for statistical analysis.
Many phenomena in the real world can be represented effectively by a form of networks. 
For example, social networks, including offline and online, can be thought of as networks.
Under an specific object of social science, some aspects of the relationship among actors 
can be modeled by networks, such as an analysis of power structure or a summary of intimacy structure on a given group. 
Another example is an application to a transmitted disease in epidemiology.
Modeling an infection using networks enables statistical analysis based on real data.

The ability to capture distinct features on an observed network enlarges statistical models' usefulness on networks. 
The statistical models give information about local relations like formation or dissolution between two actors 
as well as global features of the whole network, such as the probability that a specific structure realizes. 
The statistical models investigate the complex behavior of networks, 
including how the observed network's global features may be related to local network structure. 
During the investigation, the statistical models can capture existing dependency on a given network and provide a tool 
to understand what certain structural features are more common on rare.

Exponential-family random graph models(ERGMs) may be the most useful tools to analyze cross-sectional networks and their dependencies.
\cite{RN117}, \cite{RN123}, \cite{RN124} and \cite{RN122} proposed the models and apply them to analyze dependent structures, 
and \cite{RN115} and \cite{RN116} started the Bayesian approaches of the models.
In the ERGMs, the dependencies are captured by a series of local features called network statistics from classical graph theory. 
Since these models play an important role throughout this paper, I will introduce them more precisely in the following chapter.

The next natural question is a way to model a network sequence or network evolution over time. 
For instance, one may want to model social network structures for one year or 
simulate a disease infection based on the sequential observations in the past. 
The first exploration applying ERGMs to a network sequence was \cite{RN109}. 
After them, a notable contributions were of \cite{RN125} and \cite{RN126}, the Temporal ERGMs (TERGMs). 
In the TERGMs, the ERGMs are used by modeling the transition probability between times. 
Moreover, \cite{RN93} extend the TERGMs to the Separable Temporal ERGMs (STERGMs), 
considering the separation between duration - how long ties tend to last - and 
incidence - the rate at which new ties are formed - from 
prevalence - for example, the total number of ties - of the network properties. 
By imposing a conditional independence condition between the duration and incidence on the model, 
the models can be extended to more effectively interpretable ones than the TERGMs 
without adding much complexity.

Although the STERGMs give us strength when interpreting the fitted parameter, 
a Bayesian approach for the STERGMs has not been conducted. 
It is partly because of the hardship from the massive computation in the parameter estimation method for the models. 
Nonetheless, the Bayesian approach of STERGMs provides the statistical inference method following the standard Bayesian way, 
so the approach has a huge advantage of convenience for statistical works for fitted parameters and predictions to next time network structure. 
Thus, in this paper, I specify the Bayesian separable temporal exponential random graph models (BSTERGMs) and propose a fitting algorithm for the model parameters.
I also discuss the inference methods, the diagnostics and goodness of fit checking procedures.

Chapter 2 reviews the background materials, including the ERGMs, the BERGMs, the TERGMs, and the STERGMs. 
In chapter 3, I propose the BSTERGMs and show the detailed setting of BSTERGMs. 
In the following chapter, chapter 4, I offer the estimation method for the BSTERGMs. 
Chapter 5 explains ways to conduct diagnostics, statistical inference, and goodness of fit checks. 
Besides, I illustrate an application of BSTERGMs to network data of scholars and faculty
from the UCLA medical working project team in chapter 6, 
Finally, in chapter 7, I discuss some possible extensions and ways to supplement the BSTERGMs further for future research.

@@

\chapter{Background} \label{Chapter2}
\section{Random graphs}
Networks or random graphs play a central role as a general model 
describing relations among subjects in the statistical context. 
To begin with, I briefly summarize some terminologies and notations for concepts related to the networks.

In this paper, I will use two words, 'graphs' and 'network,' as the same meaning without distinction.
We say each subject on the data to a node. 
They sometimes are called vertex in graph theory. 
In addition, we say a tie joining two nodes to an edge. A tie can be directed or not. 
These two components have a fundamental role in describing the relational structure. 
Define a graph or a network to $y=(V,E)$, where $V$ is a set of nodes, and $E$ is a set of edges of a network.

For simplicity, I will deal with a situation without weights on each node or edge.
Then, a graph $y$ can be represented by a more straightforward form.
Assume that the number of node $n\in \mathbb{N}$ on data is given.
Then, for $\mathcal{B}=\{0,1\}$,
a set $\mathcal{Y} \subset \mathcal{B}^{n^2}$ is a set of graphs of $n$ nodes.
Now, a graph $y \in \mathcal{Y}$ can be represented by a $n \times n$ binary matrix
with elements $y_{ij}=1$ if there is a tie from $i$-th node to $j$-th node,
otherwise $y_{ij}=0$.

@@

I extend the concept of graphs to describe a situation that connections among nodes are random.
Let $\Omega$ be an event set of all possible connected situations among given nodes.
We say $Y: \Omega \to \mathcal{Y}$ is a random variable for a graph, or a random graph.
For a random graph $Y$, denote the edge between i-th node and j-th node by $Y_{ij}$ for $i,j=1,2,...,n$
satisfying $Y_{ij}=1$ if the edge is connected. Otherwise, $Y_{ij}=0$.
Note that the $Y_{ij}$s are also random variables in our context.

@@

\section{ERGMs: Exponential random graph models}
One of the natural questions from the definition of the random graphs is 
how we can model a probability distribution to the graphs.
As a answer, the Exponential random graph models (ERGMs) (\cite{RN123} and \cite{RN122})
gives a classical but flexible scheme to model for time-invariant networks represented as a single directed or undirected graph.
Moreover, the ERGMs provide a statistical approach to the systematic exploration of several local features simultaneously and 
how these local structures interact to form a global network.

Specifically, following the notation of the previous section,
let $\mathcal{Y}$ be a set of networks with $n$ nodes and $Y$ be a random graph.
Set a distribution of $Y$ to
    \[P(Y=y;\theta) = \frac{exp(\theta^{T}s(y))}{c(\theta)}\]
for some $\theta\in\mathbb{R}^p$,
where $s(y)\in\mathbb{R}^p$ is a vector which is part of sufficient network statistics of $y\in\mathcal{Y}$,
and $c(\theta)\in\mathbb{R}$ is a normalizing constant satisfying $c(\theta)=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y))$.
Models that have such a form are called the ERGMs.

@@

\section{TERGMs: Temporal Exponential family Random Graphs Models}
Let's consider a sequence of random graph $\{Y_1, Y_2, ..., Y_T\} (T\in\mathbb{N})$ and assume that
we are interested in the dynamics or evolution of the network sequence.
The Temporal Exponential family Random Graphs Models(TERGMs)
proposed by \cite{RN125} and \cite{RN126}
are models for our interest if we permit a Markov assumption
\[P(Y_2,Y_3,...,Y_T|Y_1)=P(Y_2|Y_1)P(Y_3|Y_2)...P(Y_T|Y_{T-1})\]
and focus on the transition probability along index of the sequence.
For convenience, I refer to the index $1,2,...,T$ as a time.

Specifically, let $\mathcal{Y}$ be a set of graphs with $n$ nodes. 
Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs.
Set a distribution of $Y_2,...,Y_T$ by
\[P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta) = \frac{exp(\theta^{T}s(y_t, y_{t-1}))}{c(\theta, y_{t-1})}\]
where $t=2,...,T$, $y_t\in\mathcal{Y}$, $\theta\in\mathbb{R}^p$,
sufficient network statistics $s(y_t, y_{t-1})\in\mathbb{R}^n$,
and a normalizing constant $c(\theta, y_{t-1})=\sum_{y\in\mathcal{Y}}exp(\theta^{T}s(y, y_{t-1}))$.
Models that have such a form are called the TERGMs.

@@

\section{STERGMs: Separable-Temporal Exponential family Random Graphs Models}
One solution to remedy the problems of TERGMs is 
the Separable-Temporal Exponential family Random Graphs Models, or simply STERGMs (\cite{RN93}.)
The key idea is to separate the formation models and dissolution model
for decoupling an effect on edge's duration of the model parameter from edge's incidence
when setting a distribution on a sequence of random graph.

Precisely,
let $\mathcal{Y}$ is a set of graphs with given $n\in\mathbb{N}$ nodes. 
Suppose $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T (T\in\mathbb{N})$ be random graphs.
Let $\mathcal{Y}^+|_t$ be a subset of $\mathcal{Y}$ consisting of all graphs which have equal or additional edges comparing to $y_{t-1}$.
Likewise, let $\mathcal{Y}^-|_t$ be a subset of $\mathcal{Y}$ consisting all graphs which have equal or sparse edges comparing to $y_{t-1}$.
Next, for random graphs $Y_t^+: \Omega \to\mathcal{Y}^+|_t$, $Y_t^-: \Omega \to\mathcal{Y}^-|_t$,
and graphs $y_t^+ \in \mathcal{Y}^+|_t$, $y_t^- \in \mathcal{Y}^-|_t$, set
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for parameters $\theta^+,\theta^-\in\mathbb{R}^p$, parts of sufficient network statistics $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^p$,
and normalizers $c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))$, 
$c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))$.

Then, defining operations $+,-$ on $\mathcal{Y}$ following the edgewise boolean algebra,
(For detail, Define binary operations on the edge values $\{0,1\}$ by $+$ and $-$ 
by $1+1=1, 1+0=0+1=1, 0+0=0$ and $1-1=0, 1-0=1, 0-1=0, 0-0=0$) 
and set the combined network $y_t$ to
\[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

Additionally, assume the first-order Markov assumption,
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and the conditional independence between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$.
Note that the last condition is called the separability.
Then we get
\begin{multline}
    P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-) \\
    = P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)
\end{multline}
Models that have such a form are called the STERGMs.

\section{Network Statistics for ERGMs and derived models}
Now, I briefly discuss the terms of network statistics.
As a usual, $s(y)$ of ERGMs or $s(y_{t}^.,y_{t-1})$ of TERGMs and STERGMs would 
reflect the structure of given network without identifying each node index.
Typical candidates are here.

@@

Following \cite{RN104}, usual candidates are distributions defined by some sense of the structure of a given graph. 
For a given $i, 1\leq i \leq n-1$, $D_i(y)$ is defined as the number of nodes in y whose the number of edges incident to the node equals $i$, 
and called the $i$-th order node degree distribution.

For a given $i, 1\leq i \leq n-2$, $P_i(y)$ is defined to be the number of dyads $(j,k), 1\leq j,k \leq n$
in y whose $j$ and $k$ have nodes connected by an edge each other and they share exactly $i$ connected nodes in common,
and called the $i$-th order edgewise shared partner distribution.

Note that the bundle of all order of the node degree distribution and all order of the edgewise shared partner distribution,
\[(D_1(y),D_2(y),...,D_{n-1}(y),P_1(y),P_2(y),...,P_{n-2}(y))\] 
forms sufficient statistics of a graph.
Moreover, there are relation that
\[n=D_0(y)+D_1(y)+...+D_{n-1}(y)\]
and
\[|y|=P_0(y)+P_1(y)+...+P_{n-2}(y)=\frac{1}{2}\sum_{i=1}^{n-1} iD_i(y)\]

A function of the sufficient statistics can be candidates for the model statistics.
\cite{RN128} proposed graph statistics including the $k$-star $S_k(y)$ for $1\leq k \leq n-1$ and the $k$-triangles $T_k$ for $1\leq k \leq n-2$.
The $k$-star consists of a node together with a set of $k$ of its connected nodes.
The $k$-triangle consists of $k$ triangles that share one common edge.
They can be expressed by the definition of node degree distributions and edgewise shared partner distributions,
\[S_1 = |y|, 
S_k = \sum_{i=1}^{n-1} \binom{i}{k} D_i(y), k\geq 2\]
and
\[T_1 = \frac{1}{3}\sum_{i=1}^{n-2} iP_i(y),
T_k = \sum_{i=1}^{n-2} \binom{i}{k} P_i(y), k\geq 2\]

Another interesting statistics are the geometrically weighted value of distributions defined by
\cite{RN118}, \cite{RN104}, \cite{RN128} and \cite{RN103}.
For example, the geometrically weighted (node) degree with parameter $\tau$ is defined by
\[GWD(y;\tau)=e^{\tau} \sum_{k=1}^{n-2} (1-(1-e^{-\tau})^k)D_k(y)\]
and the geometrically weighted shared partner distribution with parameter $\tau$ is defined by
\[GWESP(y;\tau)=e^{\tau} \sum_{k=1}^{n-2} (1-(1-e^{-\tau})^k)P_k(y)\]

In the context of temporal models, parameters in the model take two network as a arguments like $s(y_t,y_{t-1})$.
One of the common choices for these statistics is the bundle of sufficient statistics of first argument $y_t$.
Another common choice is using a difference, $s(y_t,y_{t-1})=s'(y_t)-s'(y_{t-1})$ where $s'$ is a network statistics of given network.
Although there are many other statistics related to the ERGMs, I omit them and end this section here.
For additional example, see \cite{RN107} for the ERGMs, and \cite{RN125}'s chapter 2.1 for the TERGMs.

@@

The goal of this chapter is to convert the STERGMs to the Bayesian setting.
With priors $p(\theta^+),p(\theta^-)$ over $\theta^+,\theta^-$,
the Bayesian model for the STERGMs can be made. I call this model the Bayesian Separable Temporal Exponential family Random Graph Models,
or the BSTERGMs.

More explicitly, let $\mathcal{Y}$ be the set of all networks with $n\in \mathbb{N}$ nodes. Let $Y_1=y_1 \in \mathcal{Y}$ be given and $Y_2,...,Y_T, T\in\mathbb{N}$ be random graphs.
As the STERGMs, define $\mathcal{Y}^+|_t$, $\mathcal{Y}^-|_t$ be subsets of $\mathcal{Y}$ consisting all graphs which have equal or additional edges comparing to $y_{t-1}$,
and which have equal or sparse edges comparing to $y_{t-1}$, respectively.
Then, for random graphs $Y_t^+: \Omega \to \mathcal{Y}^+|_t$, $Y_t^-: \Omega \to \mathcal{Y}^-|_t$,
and graphs $y_t^+ \in \mathcal{Y}^+|_t$, $y_t^- \in \mathcal{Y}^-|_t$, set formation model
\[P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) = \frac{exp((\theta^+)^{T}s(y_t^+, y_{t-1}))}{c(\theta^+, y_{t-1})}\]
and dissolution model
\[P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-) = \frac{exp((\theta^-)^{T}s(y_t^-, y_{t-1}))}{c(\theta^-, y_{t-1})}\]
for parameters $\theta^+,\theta^-\in\mathbb{R}^p$, network statistics $s(y_t^+, y_{t-1}), s(y_t^-, y_{t-1})\in\mathbb{R}^p$,
and normalizers 
\[c(\theta^+, y_{t-1})=\sum_{y^+\in\mathcal{Y}^+}exp((\theta^+)^{T}s(y^+, y_{t-1}))\]
\[c(\theta^-, y_{t-1})=\sum_{y^-\in\mathcal{Y}^-}exp((\theta^-)^{T}s(y^-, y_{t-1}))\]

Then, with the edgewise boolean algebra defined in chapter 2.5, set the combined network $y_t$ to
\[y_t=y_t^+ - (y_{t-1} - y_t^-) = y_t^- + (y_t^+ - y_{t-1})\]

and assume that the first-order Markov property
\[P(Y_2,...,Y_T|Y_1)=P(Y_2|Y_1)...P(Y_T|Y_{T-1})\]
and separability (or, conditional independence) between $Y_t^+$ and $Y_t^-$ for all $t=2,...,T$.

Then we get
\[\theta^+ \sim p(\theta^+)\]
\[\theta^- \sim p(\theta^-)\]
and
\begin{multline}
    P(Y_t=y_t|Y_{t-1}=y_{t-1};\theta^+,\theta^-) \\ 
    =P(Y_t^+=y_t^+|Y_{t-1}=y_{t-1};\theta^+) P(Y_t^-=y_t^-|Y_{t-1}=y_{t-1};\theta^-)
\end{multline}
